\paperTitle

### 2014-05-06 doc/icml2014/assumption-proof.tex
Updated assumption's proof and made directed section reflect teh same

Finally, using the tensor power method of \citep{anandkumar13tensor}, we
  get the following result on sample complexity.
\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
  \label{thm:sample-complexity}
  Wlog, let $v=1$ and $i=1$. Let $x_1, x_2, x_3$ be three views for
  $h_1$.  
  If 
\begin{align*}
  \|\hat M_{1,2} - M_{1,2}\|_{\op} &\le \epsilon & \|\hat M_{1,2,3} - M_{1,2,3}\|_{\op} &\le \epsilon,
\end{align*}
for some $\epsilon < \half$, then
with probability at least $1 - \delta$,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le  \\
    &
      O\left( k 
      \frac{{\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
      {(\sigma_{\min}(\mOpp{1}{1}) \sigma_{\min}(\mOpp{1}{2}))^{5/2}} \right) \epsilon.
\end{align*}

Furthermore, if $\sigma_{\min}(\mYpp{c}{\Pa(h_c)}) \le \sigma$ for every
  such
$\mYpp{c}{\Pa(h_c)}$, we get,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le 
      O\left( k 
      \frac{{\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
      {\sigma^{5t}} \right) \epsilon,
\end{align*}
where $t$ is the length of the topological ordering, $h_1 > \cdots
> h_t$ as defined above.
\end{theorem}
\begin{proof}
  The first statement is a result of from Theorem 5.1 of
  \citet{anandkumar13tensor}. The second statement follows directly from
  \lemmaref{mopp-singular-values}.
\end{proof}

\theoremref{sample-complexity} says that the number of samples required
  to estimate the view $\mOpp{v}{i}$ grows exponentially in the {\em
  distance} between $x_v$ and $h_i$. 
Intuitively, this is because the influence from $h_i$ to $x_v$ gets
  diluted for every latent variable it must factor through.

>>>>

Finally, using the tensor power method of \citep{anandkumar13tensor}, we
  get the following result on sample complexity.
\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
  \label{thm:sample-complexity-1}
  Wlog, let $v=1$ and $i=1$. Let $x_1, x_2, x_3$ be three views for
  $h_1$.  
  If 
\begin{align*}
  \|\hat M_{1,2} - M_{1,2}\|_{\op} &\le \epsilon & \|\hat M_{1,2,3} - M_{1,2,3}\|_{\op} &\le \epsilon,
\end{align*}
for some $\epsilon < \half$, then
with probability at least $1 - \delta$,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le  \\
    &
      O\left( k 
      \frac{{\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
      {(\sigma_{\min}(\mOpp{1}{1}) \sigma_{\min}(\mOpp{1}{2}))^{5/2}} \right) \epsilon.
\end{align*}

Furthermore, if $\sigma_{\min}(\mYpp{c}{\Pa(h_c)}) \le \sigma$ for every
  such
$\mYpp{c}{\Pa(h_c)}$, we get,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le 
      O\left( k 
      \frac{{\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
      {\sigma^{5t}} \right) \epsilon,
\end{align*}
where $t$ is the length of the topological ordering, $h_1 > \cdots
> h_t$ as defined above.
\end{theorem}
\begin{proof}
  The first statement follows directly from Theorem 5.1 of
  \citet{anandkumar13tensor} by noting that $\|M_{1,2}\|_\op \le 1$ and
  $\|M_{1,2,3}\|_\op \le 1$ (as they represent probability
  distributions). The second statement follows directly from
  \lemmaref{mopp-singular-values}.
\end{proof}

### 2014-05-06 doc/icml2014/background.tex
Updated assumption's proof and made directed section reflect teh same

\begin{assumption}
\label{asm:full-rank}
Assume the conditional moments $\mOpp{1}{1}, \mOpp{2}{1}, \mOpp{3}{1}$ have full column rank
  $k$, and that $\pi\oft 1 \succ 0$.
\end{assumption}

@@

The tensor factorization method attacks the heart of the non-convexity
  in latent-variable models.  In the next two sections,
  we will see how to build on top of this result to estimate directed (Section~\ref{sec:directed})
  and undirected (Section~\ref{sec:undirected}) models.

>>>>

\begin{assumption}
\label{asm:full-rank}
The conditional moments $\mOpp{1}{1}$, $\mOpp{2}{1}$, $\mOpp{3}{1}$ have full column rank
  $k$, and that $\pi\oft 1 \succ 0$.
\end{assumption}

@@

The tensor factorization method attacks the heart of the non-convexity
  in latent-variable models.  In the next two sections,
  we will see how to use this result to estimate directed (Section~\ref{sec:directed})
  and undirected (Section~\ref{sec:undirected}) models.

### 2014-05-06 doc/icml2014/directed.tex
Updated assumption's proof and made directed section reflect teh same

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  do not correspond directly to the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebra operations; in
  \sectionref{piecewise}, we show how we can get a more statistically
  efficient estimator using composite likelihood.

@@

\begin{lemma}[Canonical form]
  \label{lem:reduction}
Every directed (undirected) graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent (neighbour). 
There is a one-to-one correspondence between the parameters of this
  transformed model and the original one.
\end{lemma}
\begin{proof}
  \begin{figure}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
    \label{fig:reduction}
  \end{figure}

@@

\begin{lemma}[Bottlenecks guarantee exclusive views]
  \label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \cdots, h_{i_m} \} \subseteq \sG$ be
  a connected sub-graph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then they have an exclusive view.
\end{lemma}
\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not. Note that this
  proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sG'$, let $\sV_{h}$ be the set of
  views for it and let ${\sG'}^- \eqdef \sG' \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in {\sG'}^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in {\sG'}^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
  exists some $h_1, h_2 \in {\sG'}^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected to $h_2$.
  By participation in the sub-graph, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

  Thus, it must be the case that $\sV_{h} \nsubseteq \Union_{h' \neq
    h \in \sG'} \sV_{h'}$ and hence there exists a unique view for $h$.

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the sub-graph from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sG'} \sV_{h'}$
\end{proof}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE Graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE Marginals $Z_\sC$ for every clique $\sC \in \sG$

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:%\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
  every $\Re^{k \times k}$ slice of the marginals $Z_\sC$ has full column
  rank $k$ constructed by summing out the remaining indices.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the clique $\sC = \{ h_1, h_2 \}$ this
  condition implies $\pi = \Pr(h_1) \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; 
for the clique $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column rank $k$.

@@

From \citet{anandkumar13tensor} given \assumptionref{full-rank-plus},
we have that $\mOpphat{v}{i}$ converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant that
depends polynomially on the $k$-th singular value of $\mOpp{v}{i}$.

\todo{Present sample complexity bounds for $\mOpp{v}{i}$.}
\todo{Present sample complexity bounds for $Z_\sC$.}

>>>>

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  do not correspond directly to the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebra operations; in
  \sectionref{piecewise}, we show how we can get a more statistically
  efficient estimator using composite likelihoods.

@@

\begin{lemma}[Canonical form]
  \label{lem:reduction}
Every directed (undirected) graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent (neighbor). 
There is a one-to-one correspondence between the parameters of the
  transformed and original models.
\end{lemma}
\begin{proof}
  \begin{figure}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
    \label{fig:reduction}
  \end{figure}

@@

\begin{lemma}[Bottlenecks guarantee exclusive views]
  \label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \cdots, h_{i_m} \} \subseteq \sG$ be
  a connected sub-graph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then they have an exclusive view.
\end{lemma}
\begin{proof}
  The independence conditions required by the bottleneck condition
  effectively guarantee that atleast one of the three views for
  a bottleneck be exclusive. 
  \appendixref{exclusive-views} provides a complete proof.
\end{proof}

@@

\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE Graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE Marginals $Z_\sC$ for every clique $\sC \in \sG$

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of tensors on the path from $h_i$ to
  $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following
  property which we prove to be sufficient in
  \appendixref{assumption-proof}:
\begin{assumption} 
\label{asm:full-rank-plus}
The marginal distribution of every latent variable has full support:
  $\forall{h_i \in H},~ \mPi{i} \succ 0$.
For every clique $\sC \in \sG$ (including ones with observed variables),
  and every conditional distribution $\Pr(h_c \given \Pa(h_c))$, where
  $h_c \in \sC$, every {\em tensor unfolding} has full column rank, $k$.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the
  clique $\sC = \{ h_1, h_2 \}$ this condition implies $\pi = \Pr(h_1)
  \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; for the clique
  $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column
  rank $k$.

Intuitively, we need to ensure that every marginal distribution
  $\Pr(x_v)$ can be separated into $k$ different conditional distributions
  $\Pr(x_v | h_i)$. 
The full rank assumption asks that the conditional distributions are
  linearly independent; if this were not the case, the conditional
  distribution for a particular latent state could be imitated by the
  combination of several other latent states.

@@

From \citet{anandkumar13tensor} given \assumptionref{full-rank-plus}, we
  have that $\mOpphat{v}{i}$ converges to $\mOpp{v}{i}$ at a rate of
  $n^{-\frac12}$ with a constant that depends polynomially on the $k$-th
  singular value of $\mOpp{v}{i}$:

\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
  \label{thm:sample-complexity}
  Wlog, let $v=1$ and $i=1$. Let $x_1, x_2, x_3$ be three views for
  $h_1$. With probability at least $1 - \delta$,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le  \\
    &
      O\left(
      \frac{k \log(1/\delta) {\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min} } 
      {\sqrt{n}(\sigma_{\min}(\mOpp{1}{1}) \sigma_{\min}(\mOpp{1}{2}))^{5/2}} \right) \epsilon.
\end{align*}

If $\sigma_{\min}(\Pr(c \given \Pa(h_c))) \le \sigma$ for every clique
  $\sC \in \sG$ and hidden variable $h_c \in \sC$,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le 
      O\left( 
      \frac{k \log(1/\delta) {\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
      {\sqrt{n} \sigma^{5t}} \right) \epsilon,
\end{align*}
where $t$ is the length of the shortest topological ordering from $h_1$
to the unique parent of $x_1$, $h_t$.
\end{theorem}

\theoremref{sample-complexity}, proved in \appendixref{assumption-proof}
  says that the number of samples required to estimate the view
  $\mOpp{v}{i}$ grows exponentially in the {\em distance} between $x_v$
  and $h_i$. 
Intuitively, this is expected because the influence from $h_i$ to $x_v$
  gets diluted for every latent variable it must factor through.

\todo{Present sample complexity bounds for $Z_\sC$?}
